---
title: "Code-along, Week 03: Paralympics"
author: "Alex Homer"
date: "7 October 2021"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(janitor)
#For this to work, you need the `janitor` package installed.  If it isn't
#installed you'll get an error: try running
# install.packages("janitor")
#in the console, then knitting again.
library(scales)
library(RColorBrewer)

#To install the "emo" package (also necessary to knit this) run:
# devtools::install_github("hadley/emo")
#You may need to install the "devtools" package first, in the normal way.
```

## Read data
The primary data are drawn from the "Tidy Tuesday" project: [2021 Week 32: Paralympics](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-08-03/readme.md) (credit: [International Paralympic Committee](https://db.ipc-services.org/sdms/hira/web/index)).  The secondary data are from the Organisation for Economic Co-operation and Development ^[OECD (2021), Population (indicator). doi: [10.1787/d434f82b-en](https://doi.org/10.1787/d434f82b-en) (Accessed on 06 October 2021)].  The lookup tables were scraped by me from Wikipedia ^[[List of IOC country codes](https://en.wikipedia.org/wiki/List_of_IOC_country_codes) and [ISO 3166-1](https://en.wikipedia.org/wiki/ISO_3166-1).  Scraped on 6th October 2021.]

```{r read-data, message = FALSE}
paralympic_data <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-08-03/athletes.csv"
)

oecd_data <- read_csv("data/oecd_pop.csv")
ISO_lookup <- readRDS("data/ISO_lookup.rds")
IOC_lookup <- readRDS("data/IOC_lookup.rds")
```

## Cleaning up and joining data
Our data are about countries, but neither our primary nor our secondary data frame has a column which (consistently) displays the country names.  They have columns for three-letter country codes, but they are different sets of codes (International Olympic Committee codes vs. [ISO](https://en.wikipedia.org/wiki/International_Organization_for_Standardization) codes).  So we'll need to use the lookup tables to add country names.

```{r cleanup}
oecd_cleaned <- oecd_data %>%
  clean_names() %>%
  select(-c(flag_codes, indicator, frequency)) %>%
  pivot_wider(names_from = c(measure, subject), values_from = value) %>%
  rename(
    annual_growth = AGRWTH_TOT, pop_men = MLN_PER_MEN,
    pop_women = MLN_PER_WOMEN, pop_total = MLN_PER_TOT
  ) %>%
  left_join(ISO_lookup, by = c("location" = "ISO_code"))

paralympic_cleaned <- paralympic_data %>%
  select(-country) %>% #Remove incomplete country column
  left_join(IOC_lookup, by = c("abb" = "IOC_code"))
  #This creates a new country column
```

One thing we can notice is that not all the countries in the `oecd_cleaned` data yet have a name.  Let's see which.
```{r missing-countries}
oecd_cleaned %>%
  select(country, location) %>%
  distinct() %>%
  filter(is.na(country)) %>%
  select(location)
```
Seems like the remaining values are overall figures for the OECD, the world, the [G20](https://en.wikipedia.org/wiki/G20) and for what is now the [EU27](https://en.wikipedia.org/wiki/Member_state_of_the_European_Union). We'll add names, so that it doesn't get confused later with events that don't have a country name in the Paralympic data.
```{r non-countries}
oecd_cleaned <- oecd_cleaned %>%
  mutate(
    country = case_when(
      location == "OECD" ~ "Organisation for Economic Co-operation and Development",
      location == "WLD"  ~ "World",
      location == "G20"  ~ "Group of 20",
      location == "EU27" ~ "European Union"
    )
  )
```


It might also be useful to know what fraction of the world population each country represented at each year, since it allows us to control for a growing world population.  Let's do that.

```{r world-fraction}
#First get the world data out...
world_pop <- oecd_cleaned %>%
  filter(location == "WLD") %>%
  select(time, pop_total, pop_men, pop_women) %>%
  rename(
    world_pop_total = pop_total,
    world_pop_men = pop_men,
    world_pop_women = pop_women
  )

#... then add it back in using a join
oecd_cleaned <- oecd_cleaned %>%
  left_join(world_pop, by = "time") %>%
  mutate(
    frac_men = pop_men/world_pop_men,
    frac_women = pop_women/world_pop_women,
    frac_total = pop_total/world_pop_total
  ) %>%
  select(-c(world_pop_total, world_pop_men, world_pop_women))

#We could have achieved this all in one pipeline using a `right_join`, but this
#is hopefully easier to follow!
```

`r emo::ji("bulb")` *This is a useful general technique for moving summary data that's been stored in its own rows, into a new column attached to all the rows to which it relates.  Could you work out how to do the same so that, for the EU countries in the OECD, the proportion of the EU population which they represent is shown?* `r emo::ji("bulb")`

A problem remains for joining the data, though: even the country names aren't entirely the same.  Let's see which OECD countries---under these names---apparently haven't won Paralympic medals in the time frame.

```{r which-wrong-names}
paralympic_oecd_table <- paralympic_cleaned %>%
  inner_join(oecd_cleaned, by = c("country", "year" = "time")) %>%
  #Only keeps countries that have names/years in the OECD data and the Paralympic data
  count(country, name = "n_para")

oecd_countries <- oecd_cleaned %>%
  select(country) %>%
  distinct()

oecd_countries %>%
  left_join(paralympic_oecd_table, by = "country") %>%
  filter(is.na(n_para))
  #If the count is NA, that means it didn't appear in the dataset that was
  #counted, so the full_join statement has had to introduce an NA.

  #There are probably other ways to do this which may be more elegant!
```

We can now manually check these countries against the original datasets.  Of these, Costa Rica, Malta and Romania appear under those names in both lookup datasets, so they must just not appear in the Paralympic dataset.  The other five appear in the Paralympic data, but have different names.  For consistency (since not all countries appear in the OECD data), we'll keep the names from the IOC lookup table that was used for the Paralympic data ^[Yes, this means that we're referring to the United Kingdom team as "Great Britain", which only refers to three of the four nations of the UK in most other contexts.  I don't like it either, but it's pretty standard (the UK Paralympic team calls itself "ParalympicsGB"), so you'll have to take it up with the IOC!], so we'll rename some of the OECD data before joining.

```{r join}
paralympic_full <- oecd_cleaned %>%
  mutate(
    country = case_when(
      country == "Czechia"                                              ~ "Czech Republic",
      country == "Korea, Republic of"                                   ~ "South Korea",
      country == "United Kingdom of Great Britain and Northern Ireland" ~ "Great Britain",
      country == "United States of America"                             ~ "United States",
      country == "Russian Federation"                                   ~ "Russia",
      TRUE                                                              ~ country #Other cases
    )
  ) %>%
  right_join(paralympic_cleaned, by = c("country", "time" = "year")) %>%
  #Right join because we want to keep every row in the Paralympic data, and add more
  #information if available
  rename(year = time) %>% #Change back to Paralympic variables
  select(-location) #We don't need the ISO codes any more
```

This is the way to do this in one more pipeline at this stage.  In practice, after doing the investigations above to work out which country names to change, you might go back and just combine the pipelines from the start.  So it would be fine to just start with the following code to join the data, and if the purpose of this document weren't to teach you about joining data frames that's what I would have done!

## Cleaning up and joining, take two
```{r start-afresh}
oecd_cleaned <- oecd_data %>%
  clean_names() %>%
  select(-c(flag_codes, indicator, frequency)) %>%
  pivot_wider(names_from = c(measure, subject), values_from = value) %>%
  rename(
    annual_growth = AGRWTH_TOT, pop_men = MLN_PER_MEN,
    pop_women = MLN_PER_WOMEN, pop_total = MLN_PER_TOT
  ) %>%
  left_join(ISO_lookup, by = c("location" = "ISO_code")) %>%
  mutate(
    country = case_when( #Combining into one case_when
      location == "OECD" ~ "Organisation for Economic Co-operation and Development",
      location == "WLD"  ~ "World",
      location == "G20"  ~ "Group of 20",
      location == "EU27" ~ "European Union",
      location == "CZE"  ~ "Czech Republic", #Might as well use codes for all of it!
      location == "KOR"  ~ "South Korea",
      location == "GBR"  ~ "Great Britain",
      location == "USA"  ~ "United States",
      location == "RUS"  ~ "Russia",
      TRUE               ~ country #Other cases
    )
  )

world_pop <- oecd_cleaned %>%
  filter(location == "WLD") %>%
  select(time, pop_total, pop_men, pop_women) %>%
  rename(
    world_pop_total = pop_total,
    world_pop_men = pop_men,
    world_pop_women = pop_women
  )

oecd_cleaned <- oecd_cleaned %>%
  left_join(world_pop, by = "time") %>%
  mutate(
    frac_men = pop_men/world_pop_men,
    frac_women = pop_women/world_pop_women,
    frac_total = pop_total/world_pop_total
  ) %>%
  select(-c(world_pop_total, world_pop_men, world_pop_women))

paralympic_full <- paralympic_data %>%
  select(-country) %>%
  left_join(IOC_lookup, by = c("abb" = "IOC_code")) %>%
  left_join(oecd_cleaned, by = c("country", "year" = "time")) %>%
  select(-location)
```

The columns are in a different order from what we got the first time around, but it's otherwise the same data frame.  (Notice this only works because we never overwrote the originally loaded data frames after loading them.  If we had we'd have to load them again.)

## Analysis
Let's see if there's any link between winning medals in a sport and the population of your country.  The sport appears to be contained in the `type` variable.

```{r sports}
paralympic_full %>% count(type)
```

We'll use the fraction of the world population for people of a given (binary) gender ^[Some people are non-binary, of course, and so don't fit into either of these categories.  The statistics from the OECD don't reflect that, much as the Census 2021 in England and Wales asked for respondents to state a binary sex as well as giving a free text box to state their gender.  In Paralympic sports there are only two gender categories.] as a variable to standardise by.  This is going to take some manipulation.

```{r sport v population, warning = FALSE}
paralympic_full <- paralympic_full %>%
  mutate(
    frac_for_gender = case_when(
      gender == "Men"   ~ frac_men,
      gender == "Women" ~ frac_women,
      TRUE              ~ frac_total #Fallback if gender field is missing
    )
  )

paralympic_full %>%
  ggplot(aes(x = type, y = frac_for_gender)) +
  geom_violin()
```

This gives a warning that values were removed (hidden in the knitted document), because we only have the data for the OECD countries.  Also our violin plots have some very high values, which makes them hard to read.  We could rescale the axis to be logarithmic...

```{r log violin, warning = FALSE}
paralympic_full %>%
  ggplot(aes(x = type, y = frac_for_gender)) +
  geom_violin() +
  scale_y_log10()
```

... though it is quite possible then that the density is no longer accurate given a log-transformation.

What if we look at years?

```{r years}
paralympic_full %>%
  ggplot(aes(x = year, y = frac_for_gender)) +
  geom_jitter(aes(colour = type))
```

But the problem with this is that we have too many observations the same.

```{r us}
paralympic_full %>%
  select(year, frac_for_gender, country) %>%
  distinct() %>%
  ggplot(aes(x = year, y = frac_for_gender)) +
  geom_point(aes(colour = country)) +
  geom_text(
    aes(label = country),
    nudge_y = 0.005,
    check_overlap = TRUE
  )
```

Okay, let's try something else!  For each of the OECD countries, let's calculate the number of gold, silver and bronze medals they had.  First, we'll need to modify the `medal` variable, because this has a natural order that we need to reflect: Gold > Silver > Bronze.

```{r medal-colour}
paralympic_full <- paralympic_full %>%
  mutate(
    medal = factor(
      medal,
      ordered = TRUE,
      levels = c("Bronze", "Silver", "Gold")
    )
  )

paralympic_full %>%
  filter(!is.na(pop_total)) %>% #Removes non-OECD countries
  ggplot(aes(y = country, fill = medal)) +
  geom_bar()
```

This would look better sorted by which country has the most medals (for which we'll need to count them before plotting), and with the medals shown with their correct colours!  Maybe gold medals should go at the right as well.

```{r medal-colour-2}
paralympic_full <- paralympic_full %>%
  mutate(
    country = fct_rev(fct_infreq(country)), #countries in order
    medal = fct_rev(medal) #medals in reverse order
  ) 

paralympic_full %>%
  filter(!is.na(pop_total)) %>% #Removes non-OECD countries
  ggplot(aes(y = country, fill = medal)) +
  geom_bar() +
  labs(
    x = "Number of medals",
    y = "Country",
    fill = "Medal type",
    title = "The US has won the most Paralympic medals of the OECD",
    caption = "Source: IPC/OECD"
  ) +
  scale_fill_manual(values = c("#D4AF37", "#C0C0C0", "#CD7F32")) +
  theme_minimal()
```

`r emo::ji("bulb")` *There are a lot of countries, some with values too small to see (which is also causing the bars to bunch together).  How could we change this to have an "other" row?  (Look into the function `fct_lump`.)* `r emo::ji("bulb")`

Okay, let's try one more thing: number of medals per unit population.  We'll divide the medal count for each year by *that year's* population.

```{r medals-per-capita}
paralympic_full %>% 
  filter(!is.na(pop_total)) %>%
  count(country, medal, year, pop_total) %>%
  mutate(medals_per_capita = n/pop_total) %>%
  mutate(country = fct_reorder(country, medals_per_capita, sum)) %>%
  ggplot(aes(x = medals_per_capita, y = country, fill = medal)) +
  geom_col() +
  labs(
    x = "Number of medals",
    y = "Country",
    fill = "Medal type",
    title = "Iceland punches above its weight..",
    subtitle = "... with the rest of Scandinavia not far behind in rank",
    caption = "Source: IPC/OECD"
  ) +
  scale_fill_manual(values = c("#D4AF37", "#C0C0C0", "#CD7F32")) +
  theme_minimal()
```

`r emo::ji("bulb")` *Remember this is only showing OECD countries: we don't have the population data for the others at present.  Is there anything we could do about this?* `r emo::ji("bulb")`

Finally, let's try and see which sports the Scandinavian countries are good at.

```{r scandinavia}
scandi_sports <- paralympic_full %>%
  mutate(type = factor(type)) %>%
  #Make sure we don't lose any sports, even if no medals were won
  filter(country %in% c("Iceland", "Sweden", "Norway", "Denmark", "Finland")) %>%
  count(country, type) %>%
  mutate(country = fct_drop(country)) #Drop non-Scandi countries from the factor

scandi_sports %>%
  complete(country, type, fill = list(n = 0)) %>%
  pivot_wider(names_from = country, values_from = n) %>%
  rename(sport = type)
```

Hmm, this is a little hard to interpret.  Perhaps we need a chart again.
```{r scandi-bar}
scandi_sports %>%
  ggplot(aes(x = country, y = n, fill = type)) +
  geom_col(position = "fill") +
  labs(
    x = "Country",
    y = "Proportion of medals",
    fill = "Sport",
    title = "Q: Which sports are Scandi countries good at?",
    caption = "Source: IPC/OECD"
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = brewer.pal(
      n = length(levels(scandi_sports$type)), name = "Paired"
    ),
    drop = FALSE
  ) +
  theme_minimal()
```

Swimming and athletics, apparently.  But what if we add a column for all countries?

```{r sports-all}
count_all <- paralympic_full %>%
  mutate(type = factor(type)) %>%
  count(type) %>%
  mutate(country = "World")

levels_store <- levels(scandi_sports$country)

scandi_sports <- scandi_sports %>%
  mutate(country = as.character(country)) %>%
  bind_rows(count_all) %>%
  mutate(country = factor(country)) %>%
  mutate(country = fct_relevel(country, c(levels_store, "World")))

scandi_sports %>%
  ggplot(aes(x = country, y = n, fill = type)) +
  geom_col(position = "fill") +
  labs(
    x = "Country",
    y = "Proportion of medals",
    fill = "Sport",
    title = "A: The same as everyone else!",
    caption = "Source: IPC/OECD"
  ) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = brewer.pal(
      n = length(levels(scandi_sports$type)), name = "Paired"
    ),
    drop = FALSE
  ) +
  theme_minimal()
```

We see that this isn't really much different from the rest of the world!  There are just a lot of medals available in these sports.

`r emo::ji("bulb")` *How could we assess in which sports these countries are doing disproportionately well?* `r emo::ji("bulb")`
